---
title: 'STAT 412/612 Class 18: Factors with `forcats`'
author: "Kelsey Gonzalez"
date: "3/29/2021"
titlegraphic: "images/AU-Logo-on-white-small.png"
output: 
  pdf_document:
   toc: true
    number_sections:  true
  html_document: default

urlcolor: "blue"
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyhead[R]{\includegraphics{../AU-Logo-on-white-small.png}}
- \fancyhead[L]{WK2-Homework}
- \fancyfoot[L]{STAT 412/612}
- \fancyfoot[C]{\thepage}
- \fancypagestyle{plain}{\pagestyle{fancy}}
- \headheight=30pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE,
                      echo = TRUE,
                      message = FALSE, 
                      warning = FALSE)
```


# Factors with `forcats`
Learning Outcomes

- Use functions from the `forcats` package to manipulate factors in R.  

### Factors
#### Introduction
- A **factor** variable is R's way to differentiate **categorical** data, i.e., a variable where each observational/experimental unit belongs to one and only one group, category, or level out of a finite set of groups, categories, or levels.
- Being designated as a factor allows R to provide special treatment to the data
  + Hair Color could be a factor with levels: Auburn, Red, Brown, Black, Blonde, ... 
  + City sizes: Small, Medium, Large, or Megalopolis
  + Major: Data Science, Mathematics, Statistics, or Other
  
- Factors may have words as labels for the levels but they are different from a character variable such as "Brown" or "Small":
1. Factors have a fixed (usually small) number of "levels" (possible values)  
  + A character variable, e.g., "City_name", can have many different values.
2. Factors have a default ordering of the levels
  + Character variables are only ordered through alpha-numeric sequencing
  + Note the Default ordering for factors in R is alpha-numeric **but you can change it** to whatever makes sense for the problem
  + Ordering is useful when determining the order to plot something.

#### Example        
- The forcats package is part of the tidyverse so let's load the tidyverse. 
- Let's use `tribble()` to create a data frame with the average highs in DC for each month.

    ```{r, message = FALSE}
    library(tidyverse)
    dcclimate <- tribble(~month, ~avehigh,
                         ##----/---------
                         "Jan",  43.4,
                         "Feb",  47.1,
                         "Mar",  55.9,
                         "Apr",  66.6,
                         "May",  75.4,
                         "Jul",  88.4,
                         "Aux",  86.5,
                         "Sep",  79.5,
                         "Oct",  68.4,
                         "Nov",  57.9,
                         "Dec",  46.8)
    head(dcclimate)
    ```
    
- Notice Month is a character variable. What happens when we plot the data?

    ```{r}
    ggplot(dcclimate, aes(x = month, y = avehigh)) +
      geom_col()
    ```
- The months are in alphabetical order, not calendar order - "A" months come first
- It's hard to see the weather for June is missing
- The 3-letter abbreviation for August is incorrect.

- We will see how Factors (and forcats) can help us address all of these issues and more!.

#### Factors are special in that they have *two* different representations of their levels
- A categorical variable, by definition, has two or more levels (groups, or categories)
- Each level has a unique label and a unique integer value
- R shows us the labels of the levels whenever we look at the data
  + The labels can be words or numbers, e.g., ratings: "red", "yellow", "green"; or groups: "1", "2", "3"
  + The labels have a default order, but you determine if the order has any meaning, e.g., "red" is lower than "yellow"
  + Use `levels(factor_var)` to access or see the labels and the ordering of the levels
- R stores the value of the factor as integer behind the scenes using the integers from 1, 2, ... number of levels.
  + Use `as.numeric(factor_var)` or `as.integer()` to access the numeric representation as doubles or as integers

#### Be careful when working with factor data as results may be counter-intuitive at first.

- Example: Assume a treatment has four levels "2", "15", "32", and "51" and we have five observations.
- Let's create a character variable `x` for the observation values (their levels)
- We can convert it to factor `xf` using `factor()` or `readr::parse_factor()` or `as.factor()`
- We can then compare `x` and `xf`
    ```{r}
    x  <- c("51", "32", "15", "2", "32")
    xf <- parse_factor(x)
    x
    class(x)
    as.numeric(x)
    
    xf
    class(xf)
    typeof(xf)
    levels(xf)
    as.numeric(xf)
    ```

- Example of converting a character vector to a factor
    ```{r}
    as.numeric(c("Hello", "Goodbye", "Hello"))
    as.numeric(factor(c("Hello", "Goodbye", "Hello")))
    as.numeric(parse_factor(c("Hello", "Goodbye", "Hello")))
    ```
- Note the different ordering of `factor()` and `parse_factor()` - we will revisit this.

- Combining factors with `c()` *combines the integer representations*, not the levels
    ```{r}
    fac1 <- factor(c("x1", "x2", "x3", "X5"))
    fac2 <- factor(c("y1", "y2", "y3"))
    c(levels(fac1),levels(fac2)) # just the character representation of the levels
    str(c(levels(fac1),levels(fac2)))
    c(fac1, fac2)
    str(c(fac1, fac2))
    str(c(as.numeric(fac1),as.numeric(fac2)))
    str(c(as.integer(fac1),as.integer(fac2)))
    ```

#### Converting Factor Variables with numbers as levels to Numeric Variables  
- Sometimes you may get data where there are numbers coded as a factor but you want to work with the actual numbers to take advantage of their mathematical relationships
- If you are 100% sure all levels of a factor variable are numeric, you can convert the values of a factor variable to be numeric.
- Does this work? Recall our vector of 5 observations: `xf`
    ```{r}
    xf
    levels(xf)
    parse_number(levels(xf)) 
    ```
- No, it just converts the 4 levels, not the whole vector of 5 observations

- Consider using the integer representation from `as.numeric()` inside select brackets
  + *Takes advantage of recycling*
    ```{r}
    xf             # five observations of four levels
    as.numeric(xf) # see the integer representation: 1,2,3,4,2
    xf[xf]         # [xf] extracts the values using the integer representation
    levels(xf)     # get the levels
    levels(xf)[xf] # use recycling to extract the levels based on the vector's values 
    class(levels(xf)[xf]) # this is now a character vector
    ```
- Now we have a vector of characters for the numerics for all the observations in the variable
- Final step is use a readr parse function to convert
- You can do the conversion in one step as follows:
    ```{r}
    parse_number(levels(xf)[xf])
    # to confirm
    class(parse_number(levels(xf)[xf]))
    ```
  
### Creating Factors
#### readr and base R have functions to create factor variables
- Use `parse_factor()` (readr) or `factor()` (base R) to create a factor variable 
- `parse_factor()` returns better warnings, so use it whenever possible.

- Example where we try to set the levels to the standard abbreviations in calendar order

    ```{r}
    monthvec <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    
    # or consider using month.abb to eliminate typos...
    #  a built in constant for the month abbreviations
    month.abb
    monthvec <- month.abb 
   
    dcclimate %>%
      mutate(monthfc = factor(month, levels = monthvec)) ->
      dcclimate
    
    dcclimate %>%
      mutate(monthfc2 = parse_factor(month, levels = monthvec)) ->
      dcclimate
    
    dcclimate$monthfc
    as.integer(dcclimate$monthfc)
    dcclimate$monthfc2
    as.integer(dcclimate$monthfc2)
    ```

#### Working with factor levels and their order  
- If you do not specify the `levels` argument, R will assume the levels are the *unique* values of the vector.
  + `factor()` takes the order of the levels to be the same order returned by `sort()`.
  + `parse_factor()` takes the order of the levels to be the same order *as the order of the values as introduced in the data*.
  
- Example: Notice the different order of the levels 
    ```{r}
    x <- c("A", "string", "vector", "is", "a", "string", "vector")
    factor(x)
    sort(unique(x))
    parse_factor(x)
    ```

- You can always see the levels of a factor (and their order) using the `levels()` function
    ```{r}
    levels(dcclimate$monthfc)
    ```
- Other options are the `fct_unique()` and `fct_count()` functions from the forcats package.
    ```{r}
    fct_unique(dcclimate$monthfc)
    fct_count(dcclimate$monthfc)
    ```
    
- You can count the number of levels with `nlevels()`.
    ```{r}
    nlevels(dcclimate$monthfc)
    ```

- **The level order for a factor variable sets the order of the variable in the ggplot aesthetic mapping**.
- Note: `NA`s are last

    ```{r}
    ggplot(dcclimate, aes(x = monthfc, y = avehigh)) +
      geom_col()
    ```
    
- We can force missing levels to be included in the plot by using the `drop = FALSE` argument in the appropriate scale call:
  
    ```{r}
    ggplot(dcclimate, aes(x = monthfc, y = avehigh)) +
      geom_col() +
      scale_x_discrete(drop = FALSE)
    ```

### The tidyverse forcats Package

- forcats is an R package which makes two things much easier:
  + *Changing the order* of the levels of a factor variable.
  + *Changing the levels* of a factor variable.
- It also has a few other helper functions for factors.

- **All forcats functions begin with `fct_`**. 
  + You can type `fct_` and use tab-completion to scroll through the possible functions.
- forcats is a part of the tidyverse, so you don't need to load it separately after you load the tidyverse.
    
#### Changing the Order of the Levels
- We can reorder based on the values of another variable, reverse their order, or put them in a custom order.
- Let's look at the `gss_cat` data set which is a subset of the data from the 
  [General Social Survey](https://en.wikipedia.org/wiki/General_Social_Survey).
- `gss_cat` is part of the forcats package so we can load it with `data()`
- Let's use `tibble::glimpse()` as `gss_cat` has more variables than fit on the page. 
- `glimpse()` transposes the view so the variables show down the page so we can see them all, with as many observations as fit on the width of the page.
    ```{r}
    data(gss_cat)
    head(gss_cat)
    glimpse(gss_cat)
    ```
 - Note there are six factor variables in `gss_cat`

- You often want to change the order of the levels of a factor to make plots more insightful or easier to interpret.

- Let's plot the average TV hours by Religion using the current (default) order of the levels present in the data.
    ```{r}
    levels(gss_cat$relig)
    gss_cat %>%
      group_by(relig) %>%
      summarize(tvhours_mean = mean(tvhours, na.rm = TRUE)) ->
      tvdat
    
    ggplot(tvdat, aes(x = tvhours_mean, y = relig)) +
      geom_point() +
      xlab("Average TV Hours") +
      ylab(NULL)
    ```
- This plot is hard to interpret as the religions don't seem to be in any particular order and there is no clear relationship with the TV hours 
- Let's reorder the levels for our plot

- Use `fct_reorder()`  to reorder the levels of a factor *based on the values of another variable*. The arguments are:
  + `f`: The factor vector.
  + `x`: A *numeric vector* you want to use to reorder the levels (default is ascending order (`desc = FALSE`)).
  + `fun`: A function you may want to apply to `x`, the result of which will be used to order the levels of `f`.
      + default is the median
- If we want to "permanently" change the levels we put the `fct_reorder()` inside a mutate.

- Let's reorder `relig` based on the mean of TV hours we calculated earlier.  
    ```{r}
    levels(tvdat$relig)
    tvdat %>%
      mutate(relig = fct_reorder(relig, tvhours_mean)) ->
      tvdat
    levels(tvdat$relig)
    ```

- The plot now uses the new order in the data to order the values on the y-axis.
    ```{r}
    ggplot(tvdat, aes(x = tvhours_mean, y = relig)) +
      geom_point() +
      xlab("Average TV Hours") +
      ylab(NULL)
    # Note, there is no data for "Not Applicable"
    ```

- We can also reorder inside the `aes()` call of the plot without using `mutate()`.
- Let's look at the max of TV hours.
    ```{r}
    gss_cat %>% 
      group_by(relig) %>% 
      summarize(max_tv = max(tvhours, na.rm = TRUE)) %>% 
    ggplot(aes(x = max_tv, y = fct_reorder(relig, max_tv))) +
      geom_point() +
      xlab("Max TV Hours") +
      ylab(NULL)
    ```

- `fct_rev()` reverses the order of the factors.
    ```{r}
    tvdat %>%
      mutate(relig = fct_rev(relig)) %>%
      ggplot(aes(x = tvhours_mean, y = relig)) +
        geom_point() +
        xlab("Average TV Hours") +
        ylab(NULL)
    ```
    
- `fct_relevel()` allows you to *move existing levels to any location you choose*.
  + Use `after =` to set the new location
  + Any levels not mentioned will be left in existing order

- Let's look at the current levels and then several moves.
    ```{r}
    levels(tvdat$relig)
    ```

- Move "None" to first level
    ```{r}
    fct_relevel(tvdat$relig, "None") %>%
      levels()
    ```

- Move "None" to the third level
    ```{r}
    fct_relevel(tvdat$relig, "None", after = 2) %>%
      levels()
    ```

- Move "None" to the last level
    ```{r}
    fct_relevel(tvdat$relig, "None", after = nlevels(tvdat$relig)) %>%
      levels()
    ```

- `fct_relevel()` returns a warning if you try to move a level that does not exist.
- "Cthulhuism" is not a level.
    ```{r}
    fct_relevel(tvdat$relig, "Cthulhuism")
    ```

- You can move multiple levels at once using `c()`.
    ```{r}
    fct_relevel(tvdat$relig, c("Catholic", "None", "Protestant"), after = 3L) %>% 
      levels()
    ```

##### Exercises: 
1. Look at the help examples for `fct_relevel()` 
- Change the levels of the `partyid` variable so the levels are sorted in reverse alphabetical order.
  
    ```{r,  }
    gss_cat %>%
      mutate(partyid = fct_relevel(partyid, 
                                   sort(levels(partyid), decreasing = TRUE))) %>%
      count(partyid)
    ```
  
2. Move the `"Not applicable"` level for the `rincome` variable to be the next-to-last level.
- Do not use the `nlevels()` function
  
    ```{r,  }
    gss_cat %>%
      mutate(rincome = fct_relevel(rincome, "Not applicable", after = length(levels(gss_cat$rincome))-2)) %>%
      count(rincome) %>% tail()
    ```

#### Modify the Labels of the Factor Levels

- Let's look at the levels of `partyid` in `gss_cat`.

    ```{r}
    levels(gss_cat$partyid)
    ```
    
- Use `fct_recode()` (inside a `mutate()`) to *change the labels* for the levels.
- The new label level goes on the left of the equals sign. 
- The old label level goes on the right. 
- Think `New <- Old` like `mutate()` or `summarize()`

    ```{r}
    gss_cat %>%
      mutate(partyid = fct_recode(partyid,
                                  "Republican, strong"    = "Strong republican",
                                  "Republican, weak"      = "Not str republican",
                                  "Independent, near rep" = "Ind,near rep",
                                  "Independent, near dem" = "Ind,near dem",
                                  "Democrat, weak"        = "Not str democrat",
                                  "Democrat, strong"      = "Strong democrat"
                                  )) ->
      gss_cat
    levels(gss_cat$partyid)
    ```
  

##### Exercise
- Modify the factor levels of `marital` to be abbreviations of their long-names. 
- For example, "Divorced" can just be "D"

    ```{r,  }
    fct_recode(gss_cat$marital,
               "NA" = "No answer",
               "NM" = "Never married",
               "S"  = "Separated",
               "D"  = "Divorced",
               "W"  = "Widowed",
               "M"  = "Married") %>% 
      levels()
    ```

  
#### Other Useful Functions.

- `fct_c()`: is the **safe way to combine factor vectors**.

    ```{r}
    fc1 <- parse_factor(c("A", "B"))
    fc1
    fc2 <- parse_factor(c("C", "D"))
    fc2
    fct_c(fc1, fc2)
    # Not the same as
    c(fc1, fc2)
    ```

- `fct_collapse()`: combines multiple levels into one level.
- Very useful when you have man factors and you want to create a smaller set, e.g., 50 states into 4 regions

    ```{r}
    fc <- parse_factor(c("A", "B", "C", "A", "B", "C"))
    fc
    fct_collapse(fc, "AB" = c("A", "B"))
    ```

- `fct_drop()`: removes any levels that are unused.
  
    ```{r}
    fc <- parse_factor(c("A", "B"), levels = c("A", "B", "C"))
    fc
    fct_drop(fc)
    ```

- `fct_expand()`: adds one or more new levels.
- When you might want to split a factor into to new levels

    ```{r}
    fc <- parse_factor(c("A", "B"))
    fc
    fct_expand(fc, "C", "Z")
    ```

- `fct_infreq()`: Order by frequency of a level.

    ```{r}
    fc <- parse_factor(c("A", "B", "C", "B", "C", "C"))
    fct_count(fc)
    fct_infreq(fc) 
    fct_infreq(fc) %>%
      fct_count()
    ```

## References 
Wickham and Grolemund. 2016. _R for data science_ O'Reilly Media, Inc. 
* [Chapter 16](https://r4ds.had.co.nz/workflow-projects.html)
* [Chapter 15](https://r4ds.had.co.nz/r-markdown.html)
  + [Wrangling Categorical Data in R](https://doi.org/10.7287/peerj.preprints.3163v2).
  + [8.2: Chimeras of the R Inferno](https://www.burns-stat.com/pages/Tutor/R_inferno.pdf)
  + [Factors with forcats Cheat Sheet](https://github.com/rstudio/cheatsheets/blob/master/factors.pdf)
  - [Forcats Overview](https://forcats.tidyverse.org/)
  - [Dates and Times Cheat Sheet](https://github.com/rstudio/cheatsheets/blob/master/lubridate.pdf).
  - [Lubridate Overview](https://lubridate.tidyverse.org/).
