---
title: "Homework 2"
author: "Your Name"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Q1
```{r}


```

Q2
```{r}

```

#  Implement a `fizzbuzz()` function 
1. Create a new function which takes a single number as input. Use logicals and conditionals as well as the operator `%%` as required.  
- Requirements
  + If the input number is divisible by three, return `"fizz"`. 
  + If it's divisible by five, return `"buzz"`. 
  + If it's divisible by three *and* five, return `"fizzbuzz"`. 
  + Otherwise, return the input number.   
  
a. Design your function. Write out in words in a text chunk the steps your function will need to accomplish in order to convert the input number into the required text or numeric output. 
  
  1.  
  2.  
  3. 
  4. etc.*As many steps as you need*
  
b. Write R code in a code chunk to implement your steps using a variable `x`. Test it with x having values of of 3, 5, 15, and 16.


c. Once the code is working, then copy it to a new code chunk and turn it into  function with input argument of `x`. 

d. **Show your output for the following inputs**: `3`, `5`, `15`, `2`.

    ```{r}
    library(magrittr)
    fizzbuzz <- function(x) {
      is3 <- x %% 3 == 0
      is5 <- x %% 5 == 0
      if (is3 && is5) {
        return("fizzbuzz")
      } else if (is3) {
        return("fizz")
      } else if (is5) {
        return("buzz")
      } else {
        return(x)
      }
    }
    # Example why not to use print()
     fizzbuzzp <- function(x) {
      is3 <- x %% 3 == 0
      is5 <- x %% 5 == 0
      if (is3 && is5) {
        return("fizzbuzz")
      } else if (is3) {
        return("fizz")
      } else if (is5) {
        return("buzz")
      } else {
        print(x)
      }
        }
        x1 <- fizzbuzz(4)
        x2 <- fizzbuzzp(4)
```


    ```{r}
    fizzbuzz(3)
    fizzbuzz(5)
    fizzbuzz(15)
    fizzbuzz(2)
    ```

e. Update your function to include error checking  
- Ensure the input is both numeric and a single value - not a vector. 
- Test it on `cat`, and `c(1,5)`. 
- Remember, in the code chunk where you run the function, **set your code chunk parameter for error to be TRUE**, 
  + e.g. `{r, error=TRUE}`, so it will knit with the error.
    ```{r}
    fizzbuzz <- function(x) {
          stopifnot(is.numeric(x),length(x)==1)
          is3 <- x %% 3 == 0
          is5 <- x %% 5 == 0
          if (is3 && is5) {
            return("fizzbuzz")
          } else if (is3) {
            return("fizz")
          } else if (is5) {
            return("buzz")
          } else {
            return(x)
          }
        }
    ```


    ```{r, error=TRUE}
    fizzbuzz("cat")
    ```


    ```{r,error=TRUE}
    fizzbuzz(c(1,2))
    ```

f. Complete your function by inserting and completing Roxygen comments in the code chunk, above the function, to document the function.  
- Include the following elements: title, description, usage or syntax, arguments ( the params), and return value. 
```{r}

#' Check divisibility of a number 
#'
#' fizzbuz takes an input numeric value and returns the value fizz if the input is 
#' divisible by 3, buzz if the input is divisible by 5, fizzbuzz if the input is
#' divisible by 3 and 5, or else the input number.
#' usage fizzbuzz(x)
#' @param x a single numeric value.
#'
#' @return a character value of fizz, buzz, or fizzbuzz, or,
#' a numeric value of the input.
#'
#'
#' @examples 
#' fizzbuzz(3)
#' fizzbuzz(5)
#' fizzbuzz(2)
fizzbuzz <- function(x) {
      stopifnot(is.numeric(x),length(x)==1)
      is3 <- x %% 3 == 0
      is5 <- x %% 5 == 0
      if (is3 && is5) {
        return("fizzbuzz")
      } else if (is3) {
        return("fizz")
      } else if (is5) {
        return("buzz")
      } else {
        return(x)
      } # end else block
    } # end function
```

g. Create a script out of your `fizzbuzz()` function  
- Copy and paste the code from your working function into a new .R file and save in the R directory with the file name fizzbuzz_s.R
- Rename the function to fizzbuzz_s 
- Use the following code in a code chunk to show your code
  + `cat(readr::read_file("./R/fizzbuzz_s.R"))`
  + Adjust the relative path as necessary
- Write code in a new code chunk in your original homework file to source the `fizzbuzz_s()` function
- Run the function in your homework .Rmd file to show the results with the values 35, 18, 45, and -1


# Create a new `cut()` function  
1. Write a function that uses the function `cut()` to simplify this set of nested if-else statements? 
- Consider using `-Inf` and `Inf`. 
- Note, this will also output the levels of the factors.

  a. Show the output for inputs: `31`, `30`, `10`, `-10`.

    ```{r, echo = TRUE, eval = FALSE}
    if (temp <= 0) {
      "freezing"
    } else if (temp <= 10) {
      "cold"
    } else if (temp <= 20) {
      "cool"
    } else if (temp <= 30) {
      "warm"
    } else {
      "hot"
    }
    ```
    
    ```{r}
    cutright <- function(x) {
      y <- cut(x = x, 
          breaks = c(-Inf, 0, 10, 20, 30, Inf), 
          labels = c("freezing", "cold", "cool", "warm", "hot"))
      return(y)
    }

    cutright(31)
    cutright(30)
    cutright(10)
    cutright(-10)
    cutright(c(31, 30, 10, -10))
    ```
    
  b. Look at help for `cut()`. Change the call to `cut()` to handle `<` instead of `<=` in the comparisons.  
  
  c. What is the other chief advantage of the `cut()` method for this problem? 
    (Hint: what happens if you have many values in `temp`?)
    
    ```{r}
    cutleft <- function(x) {
      cut(x = x, 
          breaks = c(-Inf, 0, 10, 20, 30, Inf), 
          right = FALSE, 
          labels = c("freezing", "cold", "cool", "warm", "hot"))
    }
    ```
    
    ```{block}
    The chief advantage is the `cut()` method can handle a vector of inputs.
    For example:
    ```

    ```{r}
    cutright(c(31, 30, 10, -10))
    ```

# Using the Forward Pipe

1. Using the forward pipe `%>%`,  
- Sample from the vector `1:10` 1000 times **with replacement**, 
- Calculate the resulting sampled vector's mean, then 
- Exponentiate that mean.
   
    ```{r}
    set.seed(1)
    x <- 1:10
    x %>% 
    sample(size = 1000, replace = TRUE) %>%
      mean() %>%
      exp()
    
    # or
    sample(x=1:10, size = 1000, replace = TRUE) %>%
      mean() %>%
      exp()
    ```
   
# Calculate a proportion
- Select a random sample of 100 normally distributed values with mean 10 and variance of 3. 
- Calculate the proportion greater than 12.

    ```{r}
    set.seed(1)
    rvec <- rnorm(100, mean = 10, sd = sqrt(3))
    mean(rvec>12)
    # or
    set.seed(1)
    (rnorm(100, mean = 10, sd = sqrt(3)) > 12) %>% 
    mean()
    ```
  
# Logical Comparisons and Subsetting

- Create the values:   
  + x <- c(TRUE, FALSE, TRUE, TRUE)
  + y <- c(FALSE, FALSE, TRUE, FALSE)
  + z <- NA
- What are the results of the following:  
  + x & y
  + x & z
  + !(x | y)
  + x | y
  + y | z
  + x[y]
  + y[x]
  + x[x|y]
  
```{r}
 x <- c(TRUE, FALSE, TRUE, TRUE) 
 y <- c(FALSE, FALSE, TRUE, FALSE)
 z <- NA
 
x & y
x & z
!(x | y)
x | y
y | z
x[y]
y[x]
x[x|y]
```
